---
title: "Death causes"
output: html_notebook
bibliography: references.bib
link-citations: true
---


<style type="text/css">

body{   
    margin: 10px !important;
    padding: 10px !important;
    padding-left: 20px;
    background:#F0F0F0;
    max-width:620px;}
    
details{
    background:#e1f9fa;}

</style>

Humanity's ability to prevent death has improved tremendously in the last 150 years. Not only can we save many more sick and injured people but we also know more about preventing illness and accidents. It is for example well known that drinking and smoking increase the risk of early death while vegetables and seatbelts decrease the risk. And it is great that people know what is good and what is bad for their health, but there is a lot more to know. How much does the bad habits increase one's risk of death? And what types of illness and tragedies are related to each habit? On this website you can based on your specific habits and data, get an estimated age of death and a breakdown of how responsible each disease or habit was for your death. These numbers are computed by our mathematical model based on results from selected scientific papers. 

#### Motivating example

Imagine a world where we kept track of cause of death but there was no research in their underlying causes. Furthermore, imagine that you were an old person living in a nice house in a quiet suburbs and you asked the scientists to come up with an estimate of your risk of dying from homicide. One would think that you would have less risk of dying from homicide than, say, a drug dealer. However, the scientists in this hypothetical world would not have any information about these things. Therefore, their best estimate would be the crude rate; for example, in America the murder rate is 5.0 out of 100,000 per year, so they would tell everyone that their chance of dying from homicide was 5/100,000=0.005% each year. Imagine now that the scientists decided to actually do some research in the subject; they went out and examined how many people who were murdered in their house had a gun and compared it to how many people who weren't murdered had guns in their house. They might find that the gun-owners had approximately 1.9 times higher risk of death from homicide in their home than non-owners (as found by the researchers [@dahlberg2004guns]). If our hypothetical world had a gun ownership rate of 40% of house holds and 20% of homicides occuring in the home(like America has today), the scientists could then give 3 different answers to the question about your risk of dying from homicide:

* If they know that your household has a gun, each year the risk of you dying from homicide is $0.005\%\cdot 80\%+20\% \cdot 0.005\%\cdot \frac{1.9}{1.9\cdot 40 \%+60\%}=$ **0.00540%**
* If they know that your household doesn't have a gun, each year the risk of you dying from homicide is $0.005\%\cdot 80\%+20\% \cdot 0.005\%\cdot \frac{1}{1.9\cdot 40 \%+60\%}=$ **0.00474%**
* If they don't know whether your household has a gun, they have to use the default rate and say that each year your risk of dying from homicide is **0.005%**

The mathematical model that we use works similarly. We start with a neutral model that only knows the rates of each death cause for each age. As we put more scientific results into the model, we can more accurately estimate the risks for the individual person. Finding these results is a never-ending task and a task filled with hard decisions. We have to decide which numbers to include in the model and how to combine the results. For example, what happens to your risk of homicide if you are both a drug addict and has a gun in your home? We deal with this case by case and describe our considerations in the section [Risk Factor Choices](#risk-factor-choices).

# The model

## The neutral model

When a user haven't filled out any personal data to the model, the model will use the *neutral* model. It knows what proportion of American population dies in each 5-year age bracket every year from the [Underlying Cause of Death data set on the CDC website](https://wonder.cdc.gov/) split into ICD10 codes (which is a standardized system used by doctors to classify cause of death). There are thousands of different ICD codes so we group them into sensible categories that we call *death causes*. Using this data we can compute the  probability of dying from each death cause and expected age of death. However, these probabilities and expected lifetime will only describe the average American and not anyone in particular.

<details><summary> Formulas </summary>
For completeness, the formula that we use to compute the probability of dying from a death cause is 
$$
P(\text{die from }d)=\sum_{a=0}^{119} P(d|a) \prod_{b=0}^{a-1} \Bigl(1-\sum_{d\in D}^b P(d|a) \Bigr)
$$
where $P(d|a)$ is the probability of dying from the death cause $d$ between your $a$'th and $a+1$'th birthday. By $D$ we denote the set of all death causes. The sum goes to 119, because there is very little chance of living past 120. In the neutral model we use the proportions from the CDC as the $P(d|a)$ numbers, but when we start including risk factors we scale them using so-called risk ratios or similar.
</details>

## Including risk factors

In order to compute personalized probabilities, we use *risk factors* which are different behaviors or circumstances that influence the probability of dying. 

When the scientists in our hypothetical example computed the risk of dying from homicide, they used the following formula that we will unpack. 
$$0.005\%\cdot 80\%+20\% \cdot 0.005\%\cdot \frac{1.9}{1.9\cdot 40 \%+60\%}$$
The first part $0.005\%\cdot 80\%$ is the risk of dying from something other than a homicide in your home. For those $80\%$ of homicides we knew nothing about their underlying causes so we simply use their base rate of $0.005\%$. This means that we have effectively split up homicides into two sub death causes - homicides in your home with yearly rate $0.005\%\cdot 20\%=0.001\%$ and homicides elsewhere with yearly rate $0.004\%$. In this example the rates ($0.005\%$, $0.004\%$ and so on) are the combined rate of Americans of all ages but in the actual model we use the rates specific to a certain age.

## Normalization factor
Looking at the more complicated part of the above equation, we are left with
$$0.001\%\cdot \frac{1.9}{1.9\cdot 40 \%+60\%} $$
<details><summary>Derivation</summary>
Let us look at where it comes from. We know that the overall risk for homicides in the home is $0.001\%$ and we know that the group that has a gun in their house should have 1.9 times higher risk of being murdered. This means that if the risk in the non-owner group is $x$, the risk of the gun-owner group is $x\cdot 1.9$. The question is now what $x$ is. To find $x$ we use that the overall risk is $0.001\%$. Some of these $0.001\%$ must come from the gun owner group. If we use the fact that $40\%$ of the population owns a gun, they would contribute with $x\cdot 1.9\cdot 40\%$ of the $0.001\%$ and likewise the non-owners will contribute with $x\cdot60\%$. In total we get an equation where we can isolate $x$.

$$
x\cdot 60\%+x\cdot 1.9\cdot 40\%=0.001\%
$$
$$
x \cdot (60\%+ 1.9\cdot 40\%)=0.001\%
$$
$$
x=0.001\%\frac{1}{60\%+ 1.9\cdot 40\%}
$$

This last equation is the formula for the risk of dying from homicide in the home for non-owners. The risk in the gun-owning group is then 
$$
x\cdot 1.9=0.001\%\frac{1.9}{60\%+ 1.9\cdot 40\%}
$$
We have hereby shown the formulas.
</details>
$~$  
One way to think about the denominator $60\%+1.9\cdot 40\%=1.36$ is as the relative risk of dying from homicide in the home. This means that the average American is 1.36 times more likely to die from homicide in their home than an American who does not own a gun (because when we consider an average American there is a certain chance they have a gun in their home). Consequently, the fraction $1.9/1.36\approx1.397$ means that if you own a gun, you are 1.397 times more likely to die from homicide in your home than the average American. Because the rate $0.001\%$ applies to the average American, the product $0.001\%\cdot 1.397\approx0.0014\%$ will be the risk of dying from homicide in the home if you own a gun. Henceforth, the denominator will be called the **normalization factor**, because it scales the risks such that the overall risk is correct.

## Smoothing

Many results in health research are presented with *binned* categories. An example can be taken from @eskelinen2010caffeine, where they had examined the relation between Alzheimer's and Coffee consumption and in the their analysis they had divided people into three "bins"; people drinking 0-2 cups of coffee per day, 3-5 cups of coffee per day and 5 or more cups of coffee per day. For each category they computed an [odds ratio](#odds-ratio) that we treat as [risk ratios](#risk-ratio). Below is the input data to our program.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
source('../../Reportgeneration/DatabaseVisualization/readJsonAndPlot.R')
make_plot_for_risk_factor('Alzheimers', 'Caffeine')
```

The graph says that the risk of obtaining Alzheimers is According to these numbers an average coffee consumption of 2.49 cups of coffees per day is more than twice as dangerous as drinking 2.51 cups of coffee per day. This is not realistic and only a result of the chosen binning. To get a more realistic and intuitive model we fit a curve to binned categories.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
make_plot_for_interpolation('Alzheimers', 'Caffeine')
```

This curve is computed as a trade-off between being smooth and reflecting the original data. We have decided that the above curve could not exceed 1, because there is no indication that extreme coffee consumption leads to more Alzheimers than total coffee abstinence. 

<details> <summary>Technical details</summary>
We fit the curve by finding the natural cubic splines $\{N_i\}_{i=1, \dots, N}$ that minimizes the expression
$$
\sum_{f} \frac{1}{\sigma_f^2}\biggl(RR(f)-\frac{1}{|f|}\int_f\sum_{i=1}^n N_i(X) \text{ d}X\biggr)^2+\lambda\int_f \Bigl|\sum_{i=1}^n N_i''(X)\Bigr|^2 \text{ d}X
$$
where $f$ is a binned category, $|f|$ is the size of the bin and $\lambda$ is a tuning parameter and $\sigma_f$ is the standard deviation of the risk ratio $RR(f)$. The outer sum goes through only the finite bins and the natural cubic splines have knots around every finite bin. The tuning parameter $\lambda$ is chosen manually based on visual inspection of the fitted curve. 

For the $\sigma_f$-values we should remember that the estimated spline function will be multiplied with the inverse normalization factor in the program. This means that we are really making this fit for the normalized risk ratios and so we should set $\sigma_f$ equal to *their* standard deviations. They are unfortunately not easily computed and since we adjust the tuning parameter manually anyway, we mostly set $\sigma_f=1$. One exception is that many risk ratio tables have a *base category* which are used to compare against all other categories. This base category has, by construction, risk ratio 1 and standard deviation 0. In these cases we often constrain the spline to take that value (by using Lagrange multiplier method on the above expression). However, the normalized risk ratio of the base category does not have standard deviation zero, so the constraint does not make the computation more correct. We do it to make the resemblance between the interpolated curve and the risk ratio histogram more clear to humans.

When the input risk ratios are arranged in a multidimensional grid, we use cross products of one dimensional splines. The basis is 
$$
\{D_k(X_i)D_l(X_j)\}_{(i,j,k,l)\in I}\cup \{D_k(X_i)\}_{(k,i)\in J}
$$
where $D_k(X_i)$ is the $k$'th spline base function of the $i$'th variable. The index set $I$ includes all combinations such that $i\neq j$ and $J$ is all combinations. Let $\mathcal{N}(X)$ be the spline of the multidimensional input $X=(X_1, \dots, X_m)$. The penalization term is then
$$
\lambda \int_f \sum_{i=1}^m\sum_{j=1}^m\biggl|\frac{\partial}{\partial X_i \partial X_j}\mathcal N(X)\biggr|^2 \text{ d}X
$$

The loss function term is the same as the in the one-dimensional case except the integral is also multidimensional. 

</details>

## Interactions

Often there are much more than one risk factor for a specific death cause. An example could be that both gun ownership and living in a violent household increase your risk of dying from a homicide in the home. The question is now, what happens if one does both? In other words, what is the *interaction* between household violence and gun ownership?

<details> <summary> Homicide and physical violence in the home</summary>
#### Homicide and physical violence in the home
The researchers [@kellermann1993gun] found that the odds of dying from homicide was $4.4$ times higher if there had ever been a fight where someone was physically hurt in the home. In the same process they found that $5.7\%$ of their controls answered yes to that question. If we use that number (see our consideration under [Homicide](#homicide)) we can compute the normalization factor of $1.19=5.7\%\cdot 4.4 + 94.3\%\cdot 1$, which we can use to compute

* If you live in a home where someone has been physically hurt in a fight, the risk of dying from homicide in the home $0.001\%\cdot\frac{4.4}{1.19}=$ **0.00370%** (per year)
* If you live in a home where no one has been physically hurt in a fight, the risk of dying from homicide in the home is $0.001\%\cdot\frac{1.0}{1.19}=$ **0.00084\%** (per year)
* If you don't consider physical violence in your home, the best guess of your risk of dying from homicide in the home is the default **0.001%** (per year)
</details>

#### Combining physical violence and guns.
In the [motivating example](#motivating-example) we talked about the research from [@dahlberg2004guns] which found that a gun in one's home makes one 1.9 times more likely to be killed in the home. Other research papers have examined the same issue and they were combined in a meta analysis by [@anglemyer2014accessability]. They found that  In the program we would rather use the result from []practice we would rather use the result, while [@kellermann1993gun] found that a history of domestic physical fights increases that risk by 4.4 times. If one does both, we would want our model to say that the risk is even higher than either, but how much exactly? In this example, and in many others, we have decided that we multiply the risk ratios. That is

* If you both live in a household with physical violent fights and gun(s), you risk of dying from homicide in the home is $1.9\cdot 4.4=8.36$ times higher than if you did neither.
* If there is only physical violent fights but no guns, your risk is $4.4$ times higher than if you did neither.
* If there is only guns but no physical violence, your risk is $1.9$ times higher than if you did neither.

In this example, and in many others, multiplying them is supported by how they were computed in the original research. In [@kellermann1993gun] they computed the odds ratio $4.4$ using a method that *adjusted for gun ownership*. They assumed that odds ratios for gun ownership, physical violence, race etc. should be multiplied together and then they estimated all odds ratios at once. Consequently, $4.4$ is the most correct odds ratio when multiplied together with the gun ownership odds ratio (and all the other odds ratios in their model). That is fortunate, because that is how we use it! 

The odds ratio 1.9 from [@dahlberg2004guns] is however not adjust for physical violence. Because we want to include physical violence, this is a little drawback to using their research. In [@kellermann1993gun] they included gun ownership in their model, and that This is a drawback to using their result, but it is  scientific results that adjusts for 


Most of the mathematical models used in health research have the assumption that odds ratios or risk ratios are multiplied together. Whether the 


In other words 4.4 should be multiplied with o designed to be  4.4 is designed to be multiplied the odds ratio which makes the product of the two odds ratios most correct. rractually the odds ratio that fits their data best when it is multiplied by the odds ratio of 




# Risk factor choices

### Colon Cancer

There are many minor risk factors for colorectal cancers. So far we have included the following

* [Drinking](#colorectal-cancer-drinking). 
* Physical Activity.
* Read meat consumption.

Other relevant risk factors not included in the model are smoking, family history of this cancer, other cancers and  obesity.

#### Interactions
We assume that all the selected risk factors for colorectal cancer join multiplicatively. That is the total risk ratio 

$$\text{RR}$$


#### Colorectal cancer: Drinking
We use a analysis by @cho2004alcohol who looked a

# Glossary

### Odds ratio
An odds ratio is two odds divided by each other. In this document, an odds is a probability divided by the complementary probability. That is, if the probability is $p=0.2$, the odds is
$$
\frac{p}{1-p}=\frac{0.2}{1-0.2}=\frac{1}{4}
$$

An odds ratio is then

$$
OR=\frac{\frac{p_1}{1-p_1}}{\frac{p_2}{1-p_2}}
$$
We will almost consequently make the approximation

$$
OR=\frac{\frac{p_1}{1-p_1}}{\frac{p_2}{1-p_2}}\approx \frac{p_1}{p_2}=RR
$$
which is justified when $p_1$ and $p_2$ are small or close to each other. The RR is notation for [Risk Ratio](#risk-ratio) 

### Risk ratio

A risk ratio is two probabilities divided by each other. In this program, we will use it as 

# References
